---
title: "Exploratory Data Analysis"
author: "Lucas Avelar"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
#load libraries
library(DigitalMethodsData)
library(tidyverse)
library(ggplot2)
#load data
data("gayguides")
```
#Step 1 - Formulate the question
> What types of location were documented in the U.S. Territories, i.e., Puerto Rico, Guam, and Virgin Islands between 1965 and 1980? Calculate if each type increased or decreased in numbers over time. Lastly, demonstrate the percentage of locations that were assigned an amenity feature of "AYOR", "HOT" or included "inquire locally" in the description or address fields.

#Steps 2 and 3 - Read the data and Check the packaging
```{r}
#load libraries
library(DigitalMethodsData)
library(tidyverse)
library(ggplot2)
#load data and read it
data("gayguides")

#Check the packaging
nrow(gayguides)
ncol(gayguides)
```
#Step 4 - Run str()
```{r}
str(gayguides)
```
#Step 5 - Look at the top and the bottom of the data
```{r}
head(gayguides[, c(6:9)])
tail(gayguides[, c(6:9)])
```
#Step 6 - Check your "n"s
```{r}
table(gayguides$Year)
```
# this is helpful to check data integrity after reading it. As Peng suggests in his book, if you know the dataset should have records for all years between 1965 and 1985, this helps to make sure it really does and that the number of records per year does not seem off. For the purpose of this exercise, I find it reasonable to skip Step 7 "Validade with at least one external data source".

#Step 8 - Try the easy solution first (provide prima facie evidence for the original question)
```{r}
#let's subset gayguides to get to what truly matters for the question
gg.territories <- gayguides %>%
  filter(state == 'GU' | state == 'PR' | state == 'VI')
head(gg.territories)
tail(gg.territories)
```
```{r}
#let's group by type, state, and Year and get the count for every combination
gg.terr.type.year <- gg.territories %>%
  separate_rows(type, sep=",")%>%
  group_by(type, state, Year) %>%
  summarise(count = n())
```
#cool. but what now?
```{r}
#let's first mutate "gg.terr.type.year" to change the spelling of the mismatching values.
gg.terr.type.year <- gg.terr.type.year %>%
  mutate(type = fct_collapse(type,
                             "Hotels" = c("Hotel", "Hotels"),
                             "Cruisy Areas" = c("Cruising Areas", "Cruisy Areas")
  )
  )
```
#now I assume I can try to collapse observations with the same values by grouping by type again and summarising the count.
```{r}
gg.terr.type.count <- gg.terr.type.year %>%
  group_by(type) %>%
  summarise(count = sum(count)) 
```
#We can surely answer the first part of the question:
> What types of location were documented in the U.S. Territories, i.e., Puerto Rico, Guam, and Virgin Islands between 1965 and 1980?
R: Bars/Clubs, Baths, Cruisy Areas, Hotels, and Restaurants.

#now how to answer the second part?
> Calculate if each type increased or decreased in numbers over time.

```{r}
#use ggplot to create a plot that demonstrates the growth or decrease of each location type over time.
#make a plot for each of the territories.
#line chart, with years as x axis, count as y axis, and types for colors.
#state as facet.
ggplot(data = gg.terr.type.year, mapping = aes(x = Year, y = count, color = type)) + 
  geom_line() + labs(x = "Year", y = "count", title = "Location Types over Time") + facet_wrap(~state)
```
#now to answer the last part of the question:
>Lastly, demonstrate the percentage of locations that were assigned an amenity feature of "AYOR", "HOT" or included "inquire locally" in the description or address fields.

```{r}
gg.territories %>% filter(grepl('AYOR' | ))
```

